Simple Wind Models of Cool Gas Outflows as Viewed by Metal-Line Absorption and Emission

### 2010-10-12 Winds/papers/Theory/First/Submit/ms.tex
profx
git-svn-id: https://svn.ucolick.org/xsvn/trunk/radtrans@5268 ac11df5e-a788-c23b-a8e1-c758c323924d

\begin{abstract}
We analyze the absorption and emission-line profiles produced by a set
of simple, cool gas wind models motivated by galactic-scale outflow
observations.  
We examine outflows with a range of power-law density and velocity
profiles and a radiative-driven wind model.  
We focus on the \mgiid\ doublet and \ion{Fe}{2} UV1
multiplet at $\lambda \approx 2600$\AA, but the results are applicable
to other transitions that trace outflows (e.g.\ \ion{Na}{1},
\ion{H}{1} \lya, \ion{Si}{2}).   We implement monte carlo radiative
transfer techniques that track the propogation of scattered and
flourescent photons to generate 1D spectra and 2D spectral images.
[Should we make the codes publically available?  Upon request?]
[Mention fiducial model here?]
As expected, the \ion{Mg}{2} doublet exhibits strong, blue-shifted
absorption but one also predicts strong line-emission at roughly
the systemic velocity.  This line-emission `fills-in' the absorption, 
significantly reducing the effective
opacity near systemic. Analysis of cool gas outflows that entirely ignores this
line-emission may incorrectly infer that the gas is partially
covered, measure a significantly lower peak optical depth, and/or
conclude that gas at systemic velocity is absent   
(e.g.\ an ISM or slowly infalling component). 
The \ion{Fe}{2} resonance lines show similar behaviour, however 
these lines are connected to optically thin transitions to
fine-structure levels. As such, the \ion{Fe}{2} 
absorption lines more faithfully reproduce the intrinsic opacity
of the wind.   Together these results naturally explain the
absorption and emission-line characteristics observed for star-forming
galaies at $z<1$.
We also study a scenario purported to describe the outflows
of $z \sim 3$ Lyman break galaxies and argue this model fails to
reproduce the observations when (previously ignored) scattered photons
are properly considered.   
%Strong line-emission from metal-line
%transitions is a generic prediction of wind models, even in the
%presence of dust and anisotropic configurations.  
The surface brightness profiles of emission from metal-line
transitions offers a unique means of assessing the morphology and size
of galactic-scale winds motivating deep,
spatially-extended spectroscopic observations.

@@

Another factor that may reduce line-emission is a
non-isotropic wind model. The flux is lower, for example, if one
eliminates the backside to the wind (e.g.\ the source itself could
shadow a significant fraction of the backside).  
For our fiducial model, many of the photons scattered into our
sightline arise nearly normal to the sightline such that a biconicial
wind has significantly lower line-emission
($\S$~\ref{sec:anisotropic}). 
For the anisotropic winds explored in this paper,
the line-emission scales 
roughly as $\Omega/4\pi$ where $\Omega$ is the angular extent of the
wind\footnote{There is also a first order dependence on the actual
  orientation of the wind, with maximal emission when the backside is
  fully viewed.}.  Because we require that the wind 
points toward us, it is difficult to reduce $\Omega$ much
below $2 \pi$ and, therefore, anisotropy reduces the
emission by a factor of order unity.  
Similarly, an anisotropically emitting source would modify the predicted
line-emission.  For example, if the backside of the galaxy were
brighter/fainter then would one predict brighter/fainter emission
relative to the observed continuum.
To reduce the emission, the brightest regions
of the galaxy would need to be oriented toward Earth.  Although this
would not be a
generic orientation, most spectroscopic samples are magnitude-limited
and biased towards detecting galaxies when observed from their
brightest viewpoint.  In principle, this might imply a further
reduction of order unity.
{\bf [Put comments on biconical here]}

@@

The emission-line velocity widths also reveal
characteristics of the wind.  In particular, the \feiis\ emission-line 
widths are
sensitive to the optical depth of the wind at large velocities from
systemic.  This is illustrated in Figure~\ref{fig:obs_edelv} which plots
the 90\%\ velocity width $\Delta v_e$ of \feiic\ versus the velocity
where the wind optical depth is maximal ($v_\tau$).  We find that
$\Delta v_e$ increases with $v_\tau$ such that a large width for
\feiis\ emission requires a wind profile with large optical depths at
$\mdv < -200 \mkms$.  The broadening of \feiis\ emission occurs
because this emission is dominated by single scatterings 
which trace all components of the wind.  {\bf [Also the optical depth profile of
  FeII]}

@@

In this paper, we have explored the predicted absorption and
emission-line profiles of a set of simple galactic-scale outflow
models.  Our analysis focused on the \mgiid\ and \ion{Fe}{2} UV1
multiple of rest-frame UV transitions, but the results translate to
most other transitions used to probe cool gas.  Our primary
findings are summarized as follows:

@@

\item Isotropic, dust-free wind models conserve photon flux.
  Therefore, the blue-shifted absorption-line profiles commonly
  observed in star-forming galaxies are predicted to be accompanied by
  emission-lines with similar equivalent width.  This is even true for
  anisotropic and dust-extincted scenarios.

\item The line-emission occurs preferentially at the systemic velocity
  of the galaxy and `fills-in' the absorption profiles at velocity
  offsets $|\mdv| < 200 \mkms$ from systemic.  For transitions that
  are only coupled to the ground-state (e.g.\ \ion{Mg}{2}, \lya,
  \ion{Na}{1}). The observed absorption profiles may be significantly
  different from the intrinsic optical depth profile.

@@

\item Resonance transitions that are strongly coupled to
non-resonant lines (e.g.\ \ion{Fe}{2}, \ion{Si}{2}) offer the best
characterization of the opacity of the wind and also material 
with velocities near systemic.  The emission will generally be dominated by the
optically-thin, fine-structure transitions.

\item  Dust extinction has only modest effects 
  in models which predict only a few scatterings.  Models
  with high opacity at small radii and at systemic velocity (e.g.\ an
  optically thick, ISM component)  can effectively extinguish
  resonantly trapped emission (\ion{Mg}{2}) but have more modest
  effect on non-resonant lines (\feiis).

\item We examined two scenarios designed to mimic the wind model
  proposed by S10 for $z \sim 3$ Lyman break galaxies.  These models
  predict substantial line-emission from scattered photons that
  greatly modify the predicted line-profiles.  
  These models do not reproduce the observed line-profiles.

@@

\item The kinematics and flux ratios of the emission lines 
  constrain the wind speed, opacity, dust extinction, and XXX of the
  wind.

\item Biconical?

@@

>>>>

\begin{abstract}
We analyze the absorption and emission-line profiles produced by a set
of simple, cool gas wind models motivated by galactic-scale outflow
observations.  
We implement monte carlo radiative
transfer techniques that track the propogation of scattered and
flourescent photons to generate 1D spectra and 2D spectral images.
{\bf [Should we make the codes publically available?  Upon request?]}
%We examine outflows with a range of power-law density and velocity
%profiles and a radiative-driven wind model.  
We focus on the \mgiid\ doublet and \ion{Fe}{2} UV1
multiplet at $\lambda \approx 2600$\AA, but the results are applicable
to other transitions that trace outflows (e.g.\ \ion{Na}{1},
\ion{H}{1} \lya, \ion{Si}{2}).   
By design, the resonance transitions show lue-shifted
absorption but one also predicts strong resonance and fine-structure line-emission at roughly
the systemic velocity.  This line-emission `fills-in' the absorption
reducing the equivalent width by up to $50\%$, shifting the centroid
by tens of \kms, and severely
reducing the effective
opacity near systemic.  Analysis of cool gas outflows that ignores this
line-emission may incorrectly infer that the gas is partially
covered, measure a significantly lower peak optical depth, and/or
conclude that gas at systemic velocity is absent   
(e.g.\ an interstellar or slowly infalling component). 
Because the \ion{Fe}{2} lines
are connected by optically-thin transitions to
fine-structure levels, their profiles
more closely reproduce the intrinsic opacity
of the wind.   Together these results naturally explain the
absorption and emission-line characteristics observed for star-forming
galaxies at $z<1$.
We also study a scenario purported to describe the outflows
of $z \sim 3$ Lyman break galaxies and argue this model fails to
reproduce the observations when radiative transfer of scattered photons
is properly considered.   
%Strong line-emission from metal-line
%transitions is a generic prediction of wind models, even in the
%presence of dust and anisotropic configurations.  
Although line-emission complicates the analysis of absorption-line
profiles, the 
surface brightness profiles 
offers a unique means of assessing the morphology and size
of galactic-scale winds.  Furthermore, the kinematics and line-ratios
offer powerful diagnostics of outflows,  motivating deep,
spatially-extended spectroscopic observations.

@@

Another factor that may reduce line-emission is an
anisotropic wind model. The flux is lower, for example, if one
eliminates the backside to the wind (e.g.\ the source itself could
shadow a significant fraction of the backside).  
Similarly, 
many of the scattered photons of the fiducial 
model arise nearly normal to the sightline such that a biconicial
wind can have significantly lower line-emission
($\S$~\ref{sec:anisotropic}). 
For the anisotropic winds explored in this paper,
the line-emission scales 
roughly as $\Omega/4\pi$ where $\Omega$ is the angular extent of the
wind\footnote{There is also a first order dependence on the actual
  orientation of the wind, with maximal emission when the backside is
  fully viewed.}.  Because we require that the wind 
points toward us, it is difficult to reduce $\Omega$ much
below $2 \pi$ and, therefore, anisotropy reduces the
emission by a factor of order unity.  
Similarly, an anisotropically emitting source would modify the predicted
line-emission.  For example, if the backside of the galaxy were
brighter/fainter then would one predict brighter/fainter emission
relative to the observed continuum.
To reduce the emission, the brightest regions
of the galaxy would need to be oriented toward Earth.  Although this
would not be a
generic orientation, most spectroscopic samples are magnitude-limited
and biased towards detecting galaxies when observed at their
brightest.  In principle, this might imply a further
reduction of order unity.

@@

The emission-line velocity widths also reveal
characteristics of the wind.  In particular, the \feiis\ emission-line 
widths are
sensitive to the optical depth of the wind at large velocities from
systemic.  This is illustrated in Figure~\ref{fig:obs_edelv} which plots
the 90\%\ velocity width $\Delta v_e$ of \feiic\ versus the velocity
where the wind optical depth is maximal ($v_\tau$).  We find that
$\Delta v_e$ increases with $v_\tau$ such that a large width for
\feiis\ emission requires a wind profile with large optical depths at
$\mdv < -200 \mkms$.  The broadening of \feiis\ emission occurs
because this emission is dominated by single scatterings 
which trace all components of the wind.

@@

In this paper, we have explored the predicted absorption and
emission-line profiles of a set of simple galactic-scale outflow
models.  Our analysis focused on the \mgiid\ and \ion{Fe}{2} UV1
multiple of rest-frame UV transitions, but the results transfer to
most other lines used to probe cool gas.  Our primary
findings are summarized as follows:

@@

\item Isotropic, dust-free wind models conserve photon flux.
  Therefore, the blue-shifted absorption-line profiles commonly
  observed in star-forming galaxies are predicted to be accompanied by
  emission-lines with similar equivalent width.  This even holds for
  non-extreme anisotropic and dust-extincted scenarios.

\item The line-emission occurs preferentially at the systemic velocity
  of the galaxy and `fills-in' the absorption profiles at velocity
  offsets $|\mdv| < 200 \mkms$ from systemic.  For transitions that
  are only coupled to the ground-state (e.g.\ \ion{Mg}{2}, \lya,
  \ion{Na}{1}),  this implies much lower equivalent widths (up to
  $50\%$) and also velocity profiles that are significantly offset
  from the intrinsic optical depth profile.

@@

\item Resonance transitions that are strongly coupled to
non-resonant lines (e.g.\ \ion{Fe}{2}, \ion{Si}{2}) 
produce emission dominated by the
optically-thin, fine-structure transitions.  As such, the resonance
absorption lines offer the best
characterization of the opacity of the wind and also of material 
with velocities near systemic.

\item  Dust extinction modestly affects 
  models where resonance photons are trapped for only a few scatterings.  Models
  with high opacity at small radii and at systemic velocity (e.g.\ an
  optically thick, ISM component)  can effectively extinguish
  resonantly trapped emission (\ion{Mg}{2}) but have more modest
  effect on non-resonant lines (\feiis).

\item We examined two scenarios designed to mimic the wind model
  proposed by S10 for $z \sim 3$ Lyman break galaxies.  These models
  predict substantial line-emission from scattered photons that
  greatly modify the predicted line-profiles so that  
  these models do not reproduce the observed line-profiles.

@@

@@

\item The kinematics and flux ratios of the emission lines 
  constrain the wind speed, opacity, dust extinction, and morphology of the
  wind.
